RegressionPvNP

### 2017-09-28 RegressionPvNP.Rmd
submitting before making major changes...

includeDemo = F # demographics

@@

dysplasia.df = apply(dysplasia.df, 2, unit.var)

@@

## TODO score complexity on the seg.vals rather than merged data frame
  dysplasia.df = score.cx(patient.data, dysplasia.df)

@@

    performance.at.1se[[as.character(a)]] = subset(cv.patient$lambdas, lambda.1se == T)
    models[[as.character(a)]] = fit0
  }
  save(plots, coefs, performance.at.1se, dysplasia.df, models, file=file)
}

@@

```{r hazards, echo=F, warning=F, message=F,eval=T, fig.height=12, fig.width=12}

hazards = apply( exp(t(dysplasia.df[, rownames(coefs[[select.alpha]])]) *  coefs[[select.alpha]][,1]), 1, function(x) {
  sd(x)/mean(x)
})
ch = as.data.frame(hazards)
ch$coef = coefs[[select.alpha]][,1]
ch$stability = featureStability
ch$label = rownames(ch)

qq = quantile(ch$hazards, seq(0,1,.15))
ggplot(ch, aes(coef, hazards, col=hazards >= qq[length(qq)])) + 
  geom_point(alpha=0.5) + 
  geom_text_repel(aes(x=coef, label=ifelse(hazards >= qq[length(qq)], label, '')),show.legend=F) +
  labs(title='Top features by hazard ratio') + theme(legend.position = 'bottom') + theme(legend.position='none')
```

@@

### How do coefficients related to the raw data?

@@

@@

Based on the relationship between the coefficients and the values of the data we would expect one.  
```{r}
# positive coefficients were related to gains in progressors
gains = subset(ch, coef > 0)
# negative coefficients were related to losses in non-progressors
losses = subset(ch, coef < 0)

@@

@@

@@

# P~NP may be different across randomly selected regions as well, again less clear
wilcox.test(randomGains[,1], randomGains[,2])
summary(aov(randomGains[,1]~randomGains[,2]))

@@

@@

```{r leaveoneout, echo=T, message=F, warning=F}

info = do.call(rbind, lapply(patient.data, function(df) df$info))
pg.samp = lapply(patient.data, function(pt) {
  info = pt$info
  info$SampleSD = NA
  info$SampleMEAN = NA
  if (length(info$Samplename) > 1) {
    info$SampleSD = apply(pt$seg.vals[,info$Samplename], 2, sd)
    info$SampleMEAN =  apply(pt$seg.vals[,info$Samplename], 2, mean)
  }
  info$Prediction = NA
  info$Prediction.Dev.Resid = NA
  info$PID = unlist(lapply(info$Path.ID, function(x) unlist(strsplit(x, 'B'))[1]))
  return(info)
})

@@

    train.rows = which(rownames(dysplasia.df) %in% samples)
    training = dysplasia.df[train.rows,]
    test = as.matrix(dysplasia.df[-train.rows,])
    if (ncol(test) <= 1) next
    #if ( nrow(test) == ncol(dysplasia.df) ) test = t(test)

@@

cutoff = pROC::coords(roc, 'best')[['threshold']]
cutoff = 0.5
```

This argues that we should set the prediction probability cutoff to `r pROC::coords(roc, 'best')[['threshold']]`.  However, for now the predictions from here on are based on the default `r cutoff` value.

@@

__All predictions from here onwards are based on the `r cutoff` cutoff, not the ROC curve__

@@

  #pred <- function(rows) length(which(rows$Prediction > 0.5))/nrow(rows)
  pred <- function(rows) length(which(rows$Prediction > cutoff))

@@

ggplot(melt(global.path[,c('ratio','total','path')], measure.vars='ratio', id.vars=c('path', 'total')), aes(path, value, group=path)) + 
  geom_bar(stat='identity', fill='dodgerblue3') + 
  geom_text(aes(y=value+2,label=paste(round(value, 1), '%', sep=''))) +
  geom_text(aes(y=3,label=paste('n=', total, sep=''))) +
  ylim(0,100) + labs(title='Samples predicted by pathology', x='Pathology', y='Percent identified')
```

@@

endos$brks = cut(endos$months.before.final, include.lowest = T, ordered_result = T,
    breaks=c(0, 1, 1*12, 2*12, 4*12, 6*12, 8*12, 10*12, 12*12, 15*12))

endos$labels = gsub('\\[|\\(|\\]', '',endos$brks)
endos$labels = sub(',', '-',endos$labels)

m = endos %>% group_by(brks, labels) %>% summarise( 'pred.r'= length(which(pred > 0))/length(brks), 'n'=length((Patient)) )
ggplot(m, aes(brks, pred.r*100, group=brks)) + ylim(0,105) +
  geom_bar(stat='identity', fill='dodgerblue3') + 
  geom_text(aes(y=25, label=paste(round(pred.r*100), '%', sep='')), nudge_y=-0.02) + 
  geom_text(aes(y=pred.r*100+1.5, label=paste('n=',n, sep='')), nudge_y=0.02) + 
  scale_x_discrete(labels=m$labels) +
  labs(x='Months', y='Percent predicted', title='Predictions prior to final endoscopy', subtitle='n = number of endoscopies')

@@

  if (bin) {
    df$endo.pred = ifelse(df$pred > 0, 'Pos','Neg')
    ggplot(df, aes(months.before.final, endo.pred, group=1,color=Patient)) + facet_grid(Patient ~.) + 
      scale_x_continuous(breaks=seq(minM, maxM, by = 12)) +
      geom_line() + geom_point() +
      geom_label_repel(aes(label=max.path, fill=Patient, color=Patient), color='white', fontface='bold', size=4, label.size=0, label.r=unit(0.25, 'lines'), label.padding=unit(0.15, 'lines')) +
      labs(x='Months before diagnosis', y='Predictive Endoscopy') + 
      theme(text=element_text(size=14, face="bold"), axis.title=element_text(size=16), legend.position='none')
  } else {
  ggplot(df, aes(months.before.final, pred.ratio, group=1, color=Patient, fill=Patient)) + facet_grid(Patient ~.) +
      ylim(0,1) + geom_abline(slope=0, intercept=seq(.25,.75,.25), color='grey', linetype='dashed') +
      geom_col(position='dodge', aes(y=pred.ratio)) + geom_line() +
      scale_x_continuous(breaks=seq(minM, maxM, by = 12)) + 
      geom_label(aes(y=ifelse(pred.ratio==1, 0.75, pred.ratio), label=max.path, fill=Patient), color='white', fontface='bold', size=4, label.size=0, label.r=unit(0.25, 'lines'), label.padding=unit(0.15, 'lines'), nudge_y=0.1) +
      labs(x='Months before diagnosis', y='Sample prediction ratio') + 
      theme(text=element_text(size=14, face="bold"), axis.title=element_text(size=16), legend.position='none')
  }
}

@@

p = plot.endo( subset(endos, Patient %in% subset(me, months <= -72)$Patient), bin=T) + 
  labs(title="Progressors", subtitle=paste(72/12, "years or more follow-up"))
print.png(p, 'progressors_1.png', width=900, height=1200)
grid.arrange(p)

p1a = plot.endo( subset(endos, Patient %in% subset(me, months <= -96)$Patient), bin=T) + 
  labs(title="Progressors", subtitle=paste(96/12, "years or more follow-up"))
print.png(p1a, 'progressors_1a.png', width=600, height=800)

p1b = plot.endo( subset(endos, Patient %in% subset(me, months <= -72 & months > -96)$Patient), bin=T) + 
  labs(title="Progressors", subtitle=paste(72/12, "-",96/12,"years or more follow-up"))
print.png(p1b, 'progressors_1b.png', width=600, height=800)

p1 = plot.endo( subset(endos, Patient %in% subset(me, months <= -72)$Patient), bin=F) +
  labs(title="Progressors", subtitle=paste(72/12, "years or more follow-up"))
print.png(p1, 'progressors_1_bar.png', width=900, height=1200)

@@

```{r, echo=F, warning=F, message=F, fig.height=22, fig.width=10}
p1 = plot.endo( subset(endos, Patient %in% subset(me, months <= -48 & months > -72)$Patient), bin=T) +
  labs(title="Progressors", subtitle=paste(48/12, "-",72/12,  "years or more follow-up"))
print.png(p1, 'progressors_2.png', width=400, height=600)

p1 = plot.endo( subset(endos, Patient %in% subset(me, months <= -48 & months > -72)$Patient), bin=F) +
  labs(title="Progressors", subtitle=paste(48/12, "-",72/12,  "years or more follow-up"))
print.png(p1, 'progressors_2_bar.png', width=400, height=600)

p2 = plot.endo( subset(endos, Patient %in% subset(me, months <= -24 & months > -48)$Patient), bin=T) +
  labs(title="Progressors", subtitle=paste(24/12, "-",48/12,  "years or more follow-up"))
print.png(p2, 'progressors_3.png', width=400, height=600)

p2 = plot.endo( subset(endos, Patient %in% subset(me, months <= -24 & months > -48)$Patient), bin=F) +
  labs(title="Progressors", subtitle=paste(24/12, "-",48/12,  "years or more follow-up"))
print.png(p2, 'progressors_3_bar.png', width=400, height=600)

p3 = plot.endo( subset(endos, Patient %in% subset(me, months >= -12)$Patient), bin=T) + 
  labs(title="Progressors", subtitle=paste("1 year follow-up"))
print.png(p3, 'progressors_4.png', width=400, height=600)

p3 = plot.endo( subset(endos, Patient %in% subset(me, months >= -12)$Patient), bin=F) +
  labs(title="Progressors", subtitle=paste("1 year follow-up"))
print.png(p3, 'progressors_4_bar.png', width=400, height=600)

@@

p = plot.endo( subset(nonp.endos, Patient %in% subset(nonp.endos, pred.ratio > 0)$Patient), bin=T) + 
  labs(title="Non-Progressors", subtitle=paste("Patients with predictive endoscopies, max", abs(min(me$months)/12), 'years follow-up'))
print.png(p, 'non-progressors.png', width=600, height=800)

p1 = plot.endo( subset(nonp.endos, Patient %in% subset(nonp.endos, pred.ratio > 0)$Patient), bin=F) + 
  labs(title="Non-Progressors", subtitle=paste("Patients with predictive endoscopies, max", abs(min(me$months)/12), 'years follow-up'))
print.png(p1, 'non-progressors_bar.png', width=600, height=800)

@@

```{r validation, message=F, warning=F, echo=F, fig.height=8, fig.width=6, eval=F}

@@

p2 = ggplot(data=subset(v.predictions, Status == 'NP'), aes(Patient, pred.ratio)) + geom_col(fill='darkgreen', color='grey') + coord_flip() +
    geom_label( aes(x=np[nrow(subset(np, pred.ratio == 1))/2,'Patient'], y=0.5, 
                    label=paste(round(nrow(np[np$pred.ratio == 1,])/nrow(np),2)*100,'%',sep='')), size=8, fill='darkgreen', color='white', fontface='bold' ) +
    geom_label( aes(x='41', y=0.5, label=paste(round(nrow(np[np$pred.ratio < 1 & np$pred.ratio > 0,])/nrow(np),3)*100,'%',sep='')), size=8, fill='darkgreen', color='white', fontface='bold' ) +
    theme(legend.position='none', axis.text.x = element_text(angle = 45, hjust = 1)) + 
    labs(title='Non Progressors', x='', y='Ratio predicted samples')
if (nrow(subset(np,pred.ratio == 0)) > 0) {
  p2 = p2 +  geom_text( aes(x=subset(np,pred.ratio == 0)$Patient[1], y=0.5, label=paste(round(nrow(pn[pn$pred.ratio <= 0,])/nrow(np),3)*100,'%',sep='')), size=8, fontface='bold' ) 
}

@@

```{r, echo=F, message=F, warning=F, fig.width=8, fig.height=12, eval=F}

@@

p1 = plot.endo(subset(vpi, Patient %in% subset(sum.validation.info, Status == 'P')$Patient), bin=T) + labs(title="Validation - Progressors")
print.png(p1, 'validation_progressors.png', width=600, height=800)

print.png(plot.endo(subset(vpi, Patient %in% subset(sum.validation.info, Status == 'P')$Patient), bin=F) + labs(title="Validation - Progressors"), 'validation_progressors_bar.png', width=600, height=800)

@@

print.png(p2,'validation_Nonprogressors.png', width=600, height=800)
print.png(plot.endo(subset(vpi, Patient %in% subset(sum.validation.info, Status == 'NP')$Patient), bin=F) + labs(title="Validation - Non Progressors"),'validation_Nonprogressors_bar.png', width=600, height=800)

@@

by.path = do.call(rbind.data.frame, lapply(vpd, function(pt) {
  final.endo = subset(pt, Endoscopy.Year == max(Endoscopy.Year))
  final.endo = final.endo[nrow(final.endo),]

@@

print.png(p3, 'validation_predictions.png', width=600, height=800)

>>>>

includeDemo = T # demographics

@@

dysplasia.df = apply(dysplasia.df, 2, unit.var)
```

@@

  ## TODO score complexity on the seg.vals directly rather than merged data frame?
  dysplasia.df = score.cx(patient.data, dysplasia.df)

@@

    performance.at.1se[[as.character(a)]] = subset(cv.patient$lambdas, lambda.1se == T)
    models[[as.character(a)]] = fit0
  }
  save(plots, coefs, performance.at.1se, dysplasia.df, models, labels, file=file)
}

@@

```{r hazards, echo=F, warning=F, message=F,eval=T, fig.height=6, fig.width=6}
hazards = apply( exp(t(dysplasia.df[, rownames(coefs[[select.alpha]])]) *  coefs[[select.alpha]][,1]), 1, function(x) {
  sd(x)/mean(x)
})
ch = as.data.frame(hazards)
ch$coef = coefs[[select.alpha]][,1]
ch$stability = featureStability
ch$label = rownames(ch)

qq = quantile(ch$hazards, seq(0,1,.15))
ggplot(ch, aes(coef, hazards, col=hazards >= qq[length(qq)])) + 
  geom_point(alpha=0.8) + 
  geom_text_repel(aes(x=coef, label=ifelse(hazards >= qq[length(qq)], label, '')),show.legend=F) +
  labs(title='Top features by hazard ratio') + theme(legend.position = 'none', text = element_text(size=14, face='bold')) 
```

@@

### How do coefficients relate to the raw data?

@@

### Examine a few example regions

```{r, echo=F, eval=F}
# p53 regions
chr17 = get.chr.lengths()[17,]

samples = names(which(dysplasia.df[,'17:1-4776188'] <= -1 ) )

#patients = subset(patient.info, Samplename %in% samples)
#table(patients$Pathology)
#nrow(patients)/nrow(subset(patient.info, Patient %in% patients$Patient))

values17 = dysplasia.df[samples, grep('^17:', colnames(dysplasia.df), value=T)]

values17 = merge(values17, subset(patient.info, Samplename %in% samples , select=c('Patient','Samplename')), by.x='row.names', by.y='Samplename')
values17 = melt(values17, measure.vars = grep('^17:',colnames(values17), value=T))
colnames(values17) = c('sample','Patient','region','value')
head(values17)

values17$selected = as.integer( as.character(values17$region) == '17:1-4776188' )

loc = do.call(rbind.data.frame, strsplit(as.character(values17$region), ':|-'))
colnames(loc) = c('chr','start', 'end')
loc[] = lapply(loc[], function(x) as.numeric(as.character(x)))

values17 = cbind(values17, loc)

genes = ccgenes[grep("BRCA1|SMARCE1|STAT3|STAT5B|TP53", ccgenes$Gene.Symbol),c('Gene.Symbol','Genome.Location')]

loc = unlist(strsplit(ccgenes[which(ccgenes$Gene.Symbol == 'TP53'),'Genome.Location'], ':|-'))
genes = cbind(genes, 'start'=loc[1], 'value'=-7, 'Row.names'='', 'Patient'='', 'selected'=as.factor(0))

#loc = do.call(rbind.data.frame, strsplit(ccgenes[grep("TP53|BRCA1|SMARCE1|STAT3|STAT5B", ccgenes$Gene.Symbol),'Genome.Location'], ':|-'))
#colnames(loc) = c('chr','start','end')
genes = cbind.data.frame('chr'=as.numeric(loc[1]), 'start'=as.numeric(loc[2]), 'end'=as.numeric(loc[3]), 'Gene.Symbol'='TP53', 'selected'=as.factor(0), 'sample'='', value=-7)

values17$sample = as.factor(values17$sample)
values17$selected = as.factor(values17$selected)

arms = cbind.data.frame('x'=rbind( with(chr17, chr.cent-cent.gap*3), with(chr17, chr.cent+cent.gap*3)), 'y'=rbind(10,10), 'label'=c('p arm','q arm'))
head(values17)

p = ggplot() + xlim(1,chr17$chr.length) + 
    #geom_rect(data=values17, aes(xmin=start, xmax=end, ymin=0, ymax=value, fill=Patient), alpha=0.2, show.legend=F) +
    geom_segment(data=values17, aes(x=start, xend=end, y=value, yend=value, color=Patient), show.legend=F, size=1.5) +
    geom_segment(data=unique(subset(values17, selected == 1, select=c('chr','start','end'))), aes(x=start, xend=end, y=0, yend=0), size=2, color='red') +
    geom_point(data=genes, aes(x=start,y=value), color='red', size=3, show.legend = F) +
    geom_text_repel(data=genes, aes(x=end, y=value, label=Gene.Symbol), color='red', show.legend = F, nudge_y=-0.1) +
    geom_vline(xintercept=chr17$chr.cent, color='red', linetype='longdash') +
    geom_label(data=arms, aes(x,y, label=label), fill='white') +
    geom_hline(yintercept = -1, color='grey39', linetype='longdash') + geom_hline(yintercept = 1, color='grey39', linetype='longdash') +
    geom_label(data=cbind('x'=chr17$chr.cent/2, 'y'=0), aes(x, y), label='Losses', color='white', fill='grey39') + 
    labs(y='Segment value', x='Chr position', title='Chromosome 17', subtitle=paste(length(table(values17$Patient)), 'patients,', length(samples), 'samples')) +
  theme(text=element_text(size=16), axis.text=element_text(size=12))

png(filename='chr17.png', width=600, height=600)
print(p)
dev.off()

print(p)
```

```{r, eval=F}
# BRCA etc
regions = subset(ch, grepl('^17', label) & coef > 0)
samples = which(dysplasia.df[,regions$label[1]] > 1 ) 
samples = union(names(samples), names(which(dysplasia.df[,regions$label[2]] > 1 )) )

cols = grep(paste(regions$label, collapse='|'), colnames(dysplasia.df))

values17 = cbind(dysplasia.df[samples, c(cols-2,cols-1,cols,cols+1,cols+2)], 'labels'=labels[samples])
values17 = merge(values17, subset(patient.info, Samplename %in% samples , select=c('Hospital.Research.ID','Patient','Samplename')), by.x='row.names', by.y='Samplename')

values17 = melt(values17)

values17$selected = as.integer(values17$variable %in% ch[grepl('^17', ch$label),'label'])

loc = do.call(rbind.data.frame, strsplit(as.character(values17$variable), ':|-'))
colnames(loc) = c('chr','start', 'end')
loc[] = lapply(loc[], function(x) as.numeric(as.character(x)))

values17 = cbind(values17, loc)

genes = ccgenes[grep("BRCA1|SMARCE1|STAT3|STAT5B", ccgenes$Gene.Symbol),c('Gene.Symbol','Genome.Location')]
loc = do.call(rbind.data.frame, strsplit(genes$Genome.Location, ':|-'))
colnames(loc) = c('chr','start','end')
loc[] = lapply(loc[], function(x) as.integer(as.character(x)))

genes = cbind.data.frame(genes, loc, 'value'=as.integer(5), 'Row.names'='', 'Patient'='', 'selected'=as.factor(0))

values17$Row.names = as.factor(values17$Row.names)
values17$selected = as.factor(values17$selected)
head(values17)
ggplot(values17, aes(start, value)) + # xlim(1,chr17$chr.length) + #facet_wrap(~Patient, ncol=1) +
    geom_segment(aes(x=start, xend=end, y=value, yend=value, fill=Patient, color=Patient), show.legend=F) + 
    geom_segment(data=unique(subset(values17, selected == 1, select=c('chr','start','end'))), aes(x=start, xend=end, y=0, yend=0), size=2, color='red') +
    geom_point(data=genes, aes(x=start,y=value), color='red', size=3, show.legend = F) +
    geom_text_repel(data=genes, aes(x=end, y=value, label=Gene.Symbol), color='red', show.legend = F) +
    geom_hline(yintercept = 1, color='grey39', linetype='longdash') +
    geom_label(x=median(values17$start, na.rm=T), y=7, label='Gains', color='white', fill='grey39') + 
    labs(y='Segment value', x='chr17', title='chr17')

png(filename='chr17.png', width=800, height=1000)
print(p)
dev.off()

# FGFR2 region
chr10 = get.chr.lengths()[10,]

samples = which(dysplasia.df[,'10:121013167-125853693'] > 1)

cols = grep('^10:121013167-125853693', colnames(dysplasia.df))
values10 = cbind(dysplasia.df[samples, c(cols-1, cols, cols+1)], 'labels'=labels[samples])
values10 = merge(values10, subset(patient.info, Samplename %in% names(samples) , select=c('Hospital.Research.ID','Patient','Samplename')), by.x='row.names', by.y='Samplename')

values10 = melt(values10)

loc = do.call(rbind.data.frame, strsplit(as.character(values10$variable), ':|-'))
colnames(loc) = c('chr','start', 'end')
loc[] = lapply(loc[], function(x) as.numeric(as.character(x)))

values10 = cbind(values10, loc)

loc = unlist(strsplit(ccgenes[which(ccgenes$Gene.Symbol == 'FGFR2'),'Genome.Location'], ':|-'))
fgfr2 = cbind.data.frame('chr'=as.numeric(loc[1]), 'start'=as.numeric(loc[2]), 'end'=as.numeric(loc[3]), 'Row.names'='FGFR2', 'Patient'='', value=5)

values10$Row.names = as.factor(values10$Row.names)
head(values10)
ggplot(values10, aes(start, value, fill=Patient, color=value)) + #xlim(1,chr10$chr.length) + #facet_wrap(~Patient, ncol=1) +
    geom_segment(aes(x=start, xend=end, y=value, yend=value), show.legend = F) + 
    geom_point(data=fgfr2, aes(x=start,y=value), color='red', size=3, show.legend = F) +
    geom_text_repel(data=fgfr2, aes(x=end, y=value, label=Row.names), show.legend = F, color='red') +
    #geom_vline(xintercept=chr10$chr.cent, color='red', linetype='longdash') +
    #geom_label(x=chr10$chr.cent-chr10$cent.gap*3, y=10, label='p arm', color='white') + 
    #geom_label(x=chr10$chr.cent+chr10$cent.gap*3, y=10, label='q arm', color='white') +
    geom_hline(yintercept = 1, color='grey39', linetype='longdash') +
    #geom_label(x=chr10$chr.cent/2, y=0, label='Losses', color='white', fill='grey39') + 
    labs(y='Segment value', x='chr10', title='chr10')  + scale_color_continuous(low='red',high='green')

@@

Based on the relationship between the coefficients and the values of the data we would expect one.  
```{r}
features.binom <- function(df) {
  p.adjust(apply(df, 1, function(x) {
    binom.test(x[2], sum(x), 0.5)$p.value
  }))  
}

# positive coefficients were related to gains in progressors
gains = subset(ch, coef > 0)
# negative coefficients were related to losses in non-progressors
losses = subset(ch, coef < 0)

@@

which(features.binom(gain) < 0.05)

fisher.test(gain, simulate.p.value = T)

@@

which(features.binom(loss) < 0.05)

@@

# P~NP may be different across randomly selected regions as well, again less clear
wilcox.test(randomGains[,1], randomGains[,2])
summary(aov(randomGains[,1]~randomGains[,2]))
fisher.test(randomGains, simulate.p.value = T)

which(features.binom(randomGains) < 0.05)

@@

which(features.binom(randomLosses) < 0.05)

@@

```{r leaveoneout, echo=T, message=F, warning=F}
info = do.call(rbind, lapply(patient.data, function(df) df$info))
pg.samp = lapply(patient.data, function(pt) {
  info = pt$info
  info$SampleSD = NA
  info$SampleMEAN = NA
  if (length(info$Samplename) > 1) {
    info$SampleSD = apply(pt$seg.vals[,info$Samplename], 2, sd)
    info$SampleMEAN =  apply(pt$seg.vals[,info$Samplename], 2, mean)
  }
  info$Prediction = NA
  info$Prediction.Dev.Resid = NA
  info$PID = unlist(lapply(info$Path.ID, function(x) unlist(strsplit(x, 'B'))[1]))
  return(info)
})

@@

    train.rows = which(rownames(dysplasia.df) %in% samples)
    training = dysplasia.df[train.rows,]
    test = as.matrix(dysplasia.df[-train.rows,])
    #if (ncol(test) <= 1) next
    if ( nrow(test) == ncol(dysplasia.df) ) test = t(test)

@@

cutoff = round(pROC::coords(roc, 'best')[['threshold']], 2)
cutoff = 0.5
```

This argues that we should set the prediction probability cutoff to `r pROC::coords(roc, 'best')[['threshold']]`.

@@

__All predictions from here onwards are performed using the `r cutoff` cutoff__

@@

  #pred <- function(rows) length(which(rows$Prediction > 0.5))/nrow(rows)
  if (pt$Status[1] == 'P') {
    pred <- function(rows) length(which(rows$Prediction > cutoff))
  } else {
     pred <- function(rows) length(which(rows$Prediction < cutoff))
  }

@@

ggplot(melt(global.path[,c('ratio','total','path')], measure.vars='ratio', id.vars=c('path', 'total')), aes(path, value, group=path)) + 
  geom_bar(stat='identity', fill='dodgerblue3') + 
  geom_text(aes(y=value+2,label=paste(round(value, 1), '%', sep=''))) +
  geom_text(aes(y=3,label=paste('n=', total, sep='')), color='white') +
  ylim(0,100) + labs(title='Progressor samples predicted by pathology', x='Pathology', y='Percent identified') + 
  theme(text=element_text(size=14, face='bold'), panel.background=element_blank(), axis.line=element_line(color='black'), panel.grid.major.x=element_blank(),
        panel.grid.major.y=element_line(color='grey'))
```

@@

endos$brks = cut(endos$months.before.final, include.lowest = T, ordered_result = T,
    breaks=c(0, 1, 1*12, 2*12, 4*12, 6*12, 8*12, 10*12, 12*12, 15*12))
endos$brks = with(endos, factor(brks, rev(levels(brks)))) # invert order

endos$labels = gsub('\\[|\\(|\\]', '',endos$brks)
endos$labels = sapply(strsplit( endos$labels, ',' ), function(x) paste(sort(as.numeric(x), decreasing=T), collapse='-' ))
#endos$labels = sub(',', '-',endos$labels)

## Flip order, 0-1 comes last
m = endos %>% group_by(brks, labels) %>% summarise( 'pred.r'= length(which(pred > 0))/length(brks), 'n'=length((Patient)) )

ggplot(m, aes(brks, pred.r*100, group=brks)) + ylim(0,105) +
  geom_bar(stat='identity', fill='dodgerblue3') + 
  geom_text(aes(y=25, label=paste(round(pred.r*100), '%', sep='')), nudge_y=-0.02, color='white') + 
  geom_text(aes(y=pred.r*100+1.5, label=paste('n=',n, sep='')), nudge_y=0.02) + 
  scale_x_discrete(labels=m$labels) + 
  labs(x='Months prior to endpoint', y='Percent predicted', title='Predictions prior to final endoscopy', subtitle='n = number of endoscopies') +
  theme(text=element_text(size=14, face='bold'), axis.text.x=element_text(angle=45,hjust=1), panel.grid.major.x=element_blank(),   panel.grid.major.y=element_line(color='grey') , axis.line=element_line(color='black'), panel.background=element_blank())

@@

  if (bin) {
    df$endo.pred = ifelse(df$pred > 0, 'Pos','Neg')
    ggplot(df, aes(months.before.final, endo.pred, group=1, color=endo.pred)) + facet_grid(Patient ~.) + 
      scale_x_continuous(breaks=seq(minM, maxM, by = 12)) +
      geom_line( color='grey39') + geom_point(aes(color=endo.pred)) +
      geom_label_repel(aes(label=max.path, fill=endo.pred), color='white', fontface='bold', size=4, label.size=0, label.r=unit(0.25, 'lines'), label.padding=unit(0.15, 'lines'), show.legend=F) + 
      scale_fill_manual(values=c('red4', 'green4')) + 
      scale_color_manual(values=c('red4', 'green4'), name='Endoscopy Prediction') +
      labs(x='Months before diagnosis', y='Predictive Endoscopy') + 
      theme(text=element_text(size=14, face="bold"), axis.title=element_text(size=16), legend.position='bottom')
  } else {
  ggplot(df, aes(months.before.final, pred.ratio, group=1, color=pred.ratio, fill=pred.ratio)) + facet_grid(Patient ~.) +
      ylim(0,1) + geom_abline(slope=0, intercept=seq(.25,.75,.25), color='grey', linetype='dashed') +
      geom_col(position='dodge', aes(y=pred.ratio), color='white') +# geom_line() +
      scale_x_continuous(breaks=seq(minM, maxM, by = 12)) + 
      geom_label(aes(y=ifelse(pred.ratio==1, 0.75, pred.ratio), label=max.path, fill=pred.ratio), color='white', fontface='bold', size=4, label.size=0, label.r=unit(0.25, 'lines'), label.padding=unit(0.15, 'lines'), nudge_y=0.1) +
      labs(x='Months before diagnosis', y='Sample prediction ratio') + 
      scale_fill_continuous(low='orangered4', high='green4', name='Sample Prediction Ratio') +
      theme(text=element_text(size=14, face="bold"), axis.title=element_text(size=16), legend.position='bottom')
  }
}

@@

p = plot.endo( subset(endos, Patient %in% subset(me, months <= -72)$Patient), bin=T) + 
  labs(title="Progressors", subtitle=paste(72/12, "years or more follow-up"))
#print.png(p, 'progressors_1.png', width=900, height=1200)
grid.arrange(p)

p1a = plot.endo( subset(endos, Patient %in% subset(me, months <= -96)$Patient), bin=T) + 
  labs(title="Progressors", subtitle=paste(96/12, "years or more follow-up"))
#print.png(p1a, 'progressors_1a.png', width=600, height=800)

p1b = plot.endo( subset(endos, Patient %in% subset(me, months <= -72 & months > -96)$Patient), bin=T) + 
  labs(title="Progressors", subtitle=paste(72/12, "-",96/12,"years or more follow-up"))
#print.png(p1b, 'progressors_1b.png', width=600, height=800)

p1 = plot.endo( subset(endos, Patient %in% subset(me, months <= -72)$Patient), bin=F) +
  labs(title="Progressors", subtitle=paste(72/12, "years or more follow-up"))
#print.png(p1, 'progressors_1_bar.png', width=900, height=1200)

@@

```{r, echo=F, warning=F, message=F, fig.height=22, fig.width=10}
p1 = plot.endo( subset(endos, Patient %in% subset(me, months <= -48 & months > -72)$Patient), bin=T) +
  labs(title="Progressors", subtitle=paste(48/12, "-",72/12,  "years or more follow-up"))
#print.png(p1, 'progressors_2.png', width=400, height=600)

p1 = plot.endo( subset(endos, Patient %in% subset(me, months <= -48 & months > -72)$Patient), bin=F) +
  labs(title="Progressors", subtitle=paste(48/12, "-",72/12,  "years or more follow-up"))
#print.png(p1, 'progressors_2_bar.png', width=400, height=600)

p2 = plot.endo( subset(endos, Patient %in% subset(me, months <= -24 & months > -48)$Patient), bin=T) +
  labs(title="Progressors", subtitle=paste(24/12, "-",48/12,  "years or more follow-up"))
#print.png(p2, 'progressors_3.png', width=400, height=600)

p2 = plot.endo( subset(endos, Patient %in% subset(me, months <= -24 & months > -48)$Patient), bin=F) +
  labs(title="Progressors", subtitle=paste(24/12, "-",48/12,  "years or more follow-up"))
#print.png(p2, 'progressors_3_bar.png', width=400, height=600)

p3 = plot.endo( subset(endos, Patient %in% subset(me, months >= -12)$Patient), bin=T) + 
  labs(title="Progressors", subtitle=paste("1 year follow-up"))
#print.png(p3, 'progressors_4.png', width=400, height=600)

p3 = plot.endo( subset(endos, Patient %in% subset(me, months >= -12)$Patient), bin=F) +
  labs(title="Progressors", subtitle=paste("1 year follow-up"))
#print.png(p3, 'progressors_4_bar.png', width=400, height=600)

@@

p = plot.endo( subset(nonp.endos, Patient %in% subset(nonp.endos, pred.ratio > 0)$Patient), bin=T) + 
  labs(title="Non-Progressors", subtitle=paste("Patients with predictive endoscopies, max", abs(min(me$months)/12), 'years follow-up'))
#print.png(p, 'non-progressors.png', width=600, height=800)

p1 = plot.endo( subset(nonp.endos, Patient %in% subset(nonp.endos, pred.ratio > 0)$Patient), bin=F) + 
  labs(title="Non-Progressors", subtitle=paste("Patients with predictive endoscopies, max", abs(min(me$months)/12), 'years follow-up'))
#print.png(p1, 'non-progressors_bar.png', width=600, height=800)

@@

```{r validation, message=F, warning=F, echo=F, fig.height=8, fig.width=6, eval=T}

@@

p2 = ggplot(data=subset(v.predictions, Status == 'NP'), aes(Patient, pred.ratio)) + geom_col(fill='darkgreen', color='grey') + coord_flip() +
    geom_label( aes(x=np[nrow(subset(np, pred.ratio == 1))/2,'Patient'], y=0.5, 
                    label=paste(round(nrow(np[np$pred.ratio == 1,])/nrow(np),2)*100,'%',sep='')), size=8, fill='darkgreen', color='white', fontface='bold' ) +
    geom_label( aes(x='41', y=0.5, label=paste(round(nrow(np[np$pred.ratio < 1 & np$pred.ratio > 0,])/nrow(np),3)*100,'%',sep='')), size=8, fill='darkgreen', color='white', fontface='bold' ) +
    theme(legend.position='none', axis.text.x = element_text(angle = 45, hjust = 1)) + 
    labs(title='Non Progressors', x='', y='Ratio predicted samples')
if (nrow(subset(np,pred.ratio == 0)) > 0) {
  p2 = p2 +  geom_text( aes(x=subset(np,pred.ratio == 0)$Patient[1], y=0.5, label=paste(round(nrow(np[np$pred.ratio <= 0,])/nrow(np),3)*100,'%',sep='')), size=8, fontface='bold' ) 
}

@@

```{r, echo=F, message=F, warning=F, fig.width=8, fig.height=12, eval=T}

@@

p1 = plot.endo(subset(vpi, Patient %in% subset(sum.validation.info, Status == 'P')$Patient), bin=T) + labs(title="Validation - Progressors")
#print.png(p1, 'validation_progressors.png', width=600, height=800)

#print.png(plot.endo(subset(vpi, Patient %in% subset(sum.validation.info, Status == 'P')$Patient), bin=F) + labs(title="Validation - Progressors"), 'validation_progressors_bar.png', width=600, height=800)

@@

#print.png(p2,'validation_Nonprogressors.png', width=600, height=800)
#print.png(plot.endo(subset(vpi, Patient %in% subset(sum.validation.info, Status == 'NP')$Patient), bin=F) + labs(title="Validation - Non Progressors"),'validation_Nonprogressors_bar.png', width=600, height=800)

@@

global.by.path = do.call(rbind.data.frame,
                         lapply(vpd[subset(sum.validation.info, Status == 'P')$Hospital.Research.ID], function(pt) {
  final.endo = subset(pt, Endoscopy.Year == max(Endoscopy.Year))
  final.endo = final.endo[nrow(final.endo),]

  #pred <- function(rows) length(which(rows$Prediction > 0.5))/nrow(rows)
  if (pt$Status[1] == 'P') {
    pred <- function(rows) length(which(rows$Prediction > cutoff))
  } else {
     pred <- function(rows) length(which(rows$Prediction < cutoff))
  }

@@

#print.png(p3, 'validation_predictions.png', width=600, height=800)
